---
title: "Isolating Data"
format: html
execute: 
  echo: true
  message: false
---


```{r}
#| label: setup
library(tidyverse)
```


# dplyr

## Import `babynames` Data  

Note, this data comes from the [US Social Security](https://www.ssa.gov/oact/babynames/) website and was created with code adapted from [Hadley Wickham](https://github.com/hadley/babynames/blob/master/data-raw/names.R), but updated to reflect the top 500 names per year. 

```{r}
#| label: load-babynames 

babynames <- read_csv("babynames_500.csv")

```



## Your Turn 1

Alter the code to select just the `n` column:

```{r}
select(babynames, n)
```

## Quiz

Which of these is NOT a way to select the `name` and `n` columns together?
> Answer: `select(babynames, ends_with("n"))` will not get us `name` and `n` columns. The rest will work.

```{r}
select(babynames, -c(year, sex, prop)) #-c select every column but...
select(babynames, name:n) #x:y select a range of columns
select(babynames, starts_with("n")) #select  columns starting with...
select(babynames, ends_with("n")) #select columns ending with...
```

## Your Turn 2

Use `filter`, `babynames`, and the logical operators to find:

* All of the names where prop is greater than or equal to 0.08  
* All the years the name "Judith" appeared in the top 500 names.    

```{r}
filter(babynames, name => 0.08)
filter(babynames, name == "Judith")

filter(babynames, name == "Judith" & year == 1950)
```

## Your Turn 3

Use Boolean operators to return only the rows that contain:

* _Boys_ named Sue  
* Names that were used by exactly 15 or 16 children in 1880  
* Names that are one of Acura, Lexus, or Yugo

```{r}
filter(babynames, name == "Sue", sex == "M")
filter(babynames, (n == 15 | n == 16) & year == 1880) 
filter(babynames, name == "Acura" | name == "Lexus" | name == "Yugo")

filter(babynames, n %in% c(15:18))
```

## Help Me

What is the smallest value of n?
What is the largest?

```{r}
arrange(babynames, n) #smallest = 12
arrange(babynames, desc(n)) #largest = 99693
```

## Your Turn 4

Use `%>%` or `|>` to write a sequence of functions that: 

1. Filters babynames to just the girls that were born in 2023, *then...*  
2. Selects the `name` and `n` columns, *then...*  
3. Arranges the results so that the most popular names are near the top.

```{r}
babynames |> 
  filter(sex == "F", year == 2023) |> 
  select(name, n) |> 
  arrange(n)

```

## Your Turn 5

1. Trim `babynames` to just the rows that contain **your** `name` and choose an option for `sex`.   
2. Trim the result to just the columns that will appear in your graph (not strictly necessary, but useful practice)  
3. Plot the results as a line graph with `year` on the x axis and `prop` on the y axis
4. Update the labels and titles  
5. Save your graph as "yourname.png" to submit for your preview work.  

```{r}
babynames |>
  filter(name == "Sam", 
         sex == "M") |> 
  select(year, 
         prop) |> 
  ggplot() +
  geom_line(aes(x = year, 
                y = prop),
            color = "#e63946") +
  theme_minimal() +
  labs(x = "Year",
       y = "Proportion of US Names \nIdentified as Male",
       title = "Popularity of the Name \"Sam\" for Males")

ggsave("popularity-of-sam-graph.png", width = 6, height = 4)
```



## Preview

Creating new variables or calculating summaries of existing variables is often useful when answering research questions.  

```{r}
#| label: preview-summaries 

babynames %>% 
  filter(name == "Jordan") |> 
  group_by(year) |> 
  summarise(total_prop = sum(prop)) |> 
  ggplot() +
    geom_line(mapping = aes(x = year, y = total_prop), size = 1) +
  labs(x = "Year", 
       y = "Proportion of US Births that Year",
       title = "Popularity of \"Jordan\" in the US Since 1880") +
  theme_light()
```





***

# Take aways

* Extract variables with `select()`  
* Extract cases with `filter()`  
* Arrange cases, with `arrange()`  


